---
title: "Data management for the Parental HIV data set from PMA6"
output: 
  html_document: 
    highlight: tango
    theme: flatly
---

# Purpose
The purpose of this file is to clean and recode the parental data set. Each time this file is run it will write a new data set to the hard drive with the current date. 

Each time a new recode or edit is added a note should be written about when and why this edit was made, and this file should be compiled. 

# Import the raw data
The raw data from the textbook is stored in the same directory as this code file. I read it in and  store it as an object `raw`, so I can use it to confirm my recodes and edits work correctly. 

```{r, message=FALSE, warning=FALSE}
library(dplyr)
raw <- read.table("Parhiv.txt", sep="\t", header=TRUE)
hiv <- raw # this is the table that will be edited
```

I personally always set the names to lower case to make it easier to type. 
```{r}
names(hiv) <- tolower(names(hiv))
```

# Examine data 
```{r}
glimpse(hiv)
```

The first thing I notice is that all the data is entered in as numeric. The codebook indicates that many of these variables are factors, so I know I'll need to convert those to factors

## Skim codebook for binary indicator changes
Sometimes programs or individuals will enter in binary data as 1/2 (no/yes) instead of 0/1 which is needed for statistical analysis. Let's handle those first. And i'll leave them as binary 0/1 instead of adding factor levels to them. Let the analyst do that. 

```{r}
hiv$siblings <- hiv$siblings -1 
hiv$school <- hiv$school -1 
hiv$hookey <- hiv$hookey -1 
```


## Change factor variables from numeric to categorical.

```{r}
hiv$gender  <- factor(hiv$gender, labels=c("Male", "Female"))
hiv$livwith <- factor(hiv$livwith, labels=c("Both parents", "One parent", "Other"))
hiv$jobmo   <- factor(hiv$jobmo, labels=c("Employed", "Unemployed", "Retired/Disabled"))
hiv$edumo   <- factor(hiv$edumo, labels=c("<HS", "HS/GED", "HS+"))
hiv$howrel  <- factor(hiv$howrel, labels=c("Very", "Somewhat", "Not at all"))
hiv$finsit  <- factor(hiv$finsit, labels=c("Very poor", "Poor", "Necessities" ,"Comfortable"))
hiv$ethn    <- factor(hiv$ethn, labels=c("Latinx", "Black", "Other"))
hiv$attserv <- factor(hiv$attserv, labels=c("Never", "Sometimes", "Frequently"))
hiv$frnds   <- factor(hiv$frnds, labels=c("good friends", "Lots acquaint.", "Few acquaint.", "not close to anyone"))
hiv$likesch <- factor(hiv$likesch, labels=c("Very much", "somewhat", "meh", "dislike somewhat", "dislike very much"))
```

Confirm that these recodes work (not showing all)
```{r}
table(raw$HOWREL, hiv$howrel, useNA="always")
```


There are 11 questions about the neighborhood that all have the same factor levels. Let's convert all of them to factors at the same time using the `across()` function from `dplyr`. 
```{r}
table(hiv$nghb1)
hiv <- hiv %>%
   mutate(across(contains("nghb"), 
                 ~factor(.x, labels = c("Not one", "Somewhat", 
                                        "Quite", "Very serious"))))
   
table(raw$NGHB1, hiv$nghb1, useNA="always")
```

### Set missing values

If haven't started smoking/drinking, then set age started to missing. 
```{r}
hiv$agesmoke[hiv$agesmoke==0] <- NA
hiv$agealc[hiv$agealc==0] <- NA
hiv$agemar[hiv$agemar==0] <- NA
```

Check for out of range values
```{r}
hiv %>% select(contains("age")) %>% summary()
```

Ages of 2 and 4 for age when they started drinking or smoking is a bit implausible. How often does that occur? Could it be a typo?

```{r}
hiv %>% 
  filter(agesmoke < 10 | agealc < 10 | agemar < 10) %>%
  select(agesmoke, agealc, agemar)
```

Too many to be typos. The 2 and perhaps the 4 are the only ones that really seem out of the ordinary. However I'll let the analyst decide to remove those at the time of analysis. 


###  Update the Parental HIV data set by creating all the subscales listed at the bottom of the codebook.
    
I will use this space to show many different ways to approach this task. The methods are not numbered by any real characteristic. For each scale in the list I just thought of a method that would work easily for the variable at hand, that I had not already demonstrated. Some methods are more advanced than others. You will find the method(s) that work best for you. 

#### Method 1: Reverse code all sub items using `recode` that require it by making new variables. Then apply the mean function across that matrix (`cbind`) of variables. 

```{r}
rpb02 <- recode(hiv$pb02, `1`=4,  `2`=3, `3`=2, `4`=1) 
table(rpb02, hiv$pb02, useNA="always") 
rpb04 <- 5-hiv$pb04 #easier way to flip a scale
rpb14 <- 5-hiv$pb14
rpb16 <- 5-hiv$pb16
rpb18 <- 5-hiv$pb18
rpb24 <- 5-hiv$pb24

hiv$parent_care <- apply(cbind(hiv$pb01, rpb02, rpb04, hiv$pb05, hiv$pb06, hiv$pb11, hiv$pb12, 
                          rpb14, rpb16, hiv$pb17, rpb18, rpb24), 1, mean)
# confirm recode
summary(hiv$parent_care)
```

#### Method 2: Reverse code sub-items at the time of combining (`cbind`). 
```{r}
hiv$parent_overprotection <- apply(cbind(5-hiv$pb03, 5-hiv$pb07, hiv$pb08, hiv$pb09, hiv$pb10, 
                                         hiv$pb13, 5-hiv$pb15, hiv$pb19, hiv$pb20, 5-hiv$pb21, 
                                         5-hiv$pb22, hiv$pb23, 5-hiv$pb25), 1, mean)
summary(hiv$parent_overprotection)
```

#### Method 3: Take the row-wise mean across relevant columns
```{r}
hiv$bsi_overall <- hiv %>% select(bsi01:bsi53) %>% rowMeans()
# confirm that some numbers were created and that the variable is not fully missing, 
# and all values are in the appropriate range of 0 to 4. 
summary(hiv$bsi_overall) 
```


#### Method 4: Using `rowMeans` function on only `selected` columns
```{r}
hiv$bsi_somat <- hiv %>% select(bsi02, bsi07, bsi23, bsi29, bsi30, bsi33 ,bsi37) %>% rowMeans()
summary(hiv$bsi_somat) 
```

#### Method 5: Using the `mutate` function, `rowwise`.

I'm doing the rest this way just because I only have to copy the mutate statement. 

```{r}
hiv <- hiv %>% 
       rowwise() %>% 
       mutate(bsi_obcomp   = mean(c(bsi05, bsi15, bsi26, bsi27, bsi32, bsi36)),
              bsi_interp   = mean(c(bsi02, bsi07, bsi23, bsi29, bsi30, bsi33 ,bsi37)),
              bsi_depress  = mean(c(bsi09, bsi16, bsi17, bsi18 ,bsi35, bsi50)),
              bsi_anxiety  = mean(c(bsi01, bsi12, bsi19, bsi38 ,bsi45, bsi49)),
              bsi_hostil   = mean(c(bsi06 ,bsi13, bsi40, bsi41, bsi46)),
              bsi_phobic   = mean(c(bsi08, bsi28, bsi31, bsi43, bsi47)),
              bsi_paranoid = mean(c(bsi04 ,bsi10 ,bsi24, bsi48 ,bsi51)),
              bsi_psycho   = mean(c(bsi03, bsi14, bsi34, bsi44, bsi53)))

hiv %>% select(contains("bsi_")) %>% summary()
```


```{r}
names(hiv)
```
See, all the subscales have been appended to the end of the data set. 


# Final glimpse
```{r}
glimpse(hiv)
```

# Save the cleaned data as a .Rdata file

```{r}
save(hiv, file = "parhiv_clean.Rata")
```

# Session Info
This document was compiled on `r Sys.time()` and with the following system information:
```{r}
sessionInfo()
```

